\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex
Pass at missingness

A varying degree of missingness is present in the genetic data, that is, for a
given sample, a subset of nucleotides may be coded as `unknown' at a set of positions.
As a result, we had to consider missingness in our implementation of the LS
Viterbi algorithm. One approach would be to `impute' all missing data
beforehand using external software to hard-call or define a distribution over
the possible missing nucleotides and use that in the determination of path
probabilities. Another, which we carry out within \texttt{sc2ts}, is to assume sites with missing nucleotides are uninformative to the path probability, and impute them as each focal sequence is analysed, on the fly. We achieve this by setting the emission probability from any state $(A,C,G,U,-)$ to `unknown' as 1. Then, after determining a most likely copying path, we can impute these missing data by simply copying nucleotides present on the most likely copying path at the sites of missing data in the focal sequence. Note that by iteratively imputing missing data in focal sequences in this way, the reference panel being used for any newly observed focal sequence is guaranteed to be devoid of missingness.

>>>>

% A varying degree of missingness is present in the genetic data, that is, for a
% given sample, a subset of nucleotides may be coded as `unknown' at a set of
% positions. As a result, we had to consider missingness in our implementation of
% the LS Viterbi algorithm.

% To deal with this missingness, we could explicitly model the missingness
% structure and generative process, or impute the missing data and revert back to
% our model without missingness. We choose the latter. To do this, we estimate a
% most likely path through the data assuming that missing nucleotides in the
% focal sequence are uninformative to the probability of any given path through
% the reference data. We then simply fill in the missing data in the focal
% sequence by copying it from the relevant positions in the most likely path. To
% ensure that missing data is not considered in determining the most likely
% copying path, we set the emission probability from any state $(A,C,G,U,-)$ to
% `unknown' as 1 (though we could have chosen any constant). As a result,
% emissions of unknown nucleotides will not contribute to differences in path
% probabilities. Note that by iteratively imputing missing data in focal
% sequences in this way, the reference panel being used for any newly observed
% focal sequence is guaranteed to be devoid of missingness.

% One approach would be to `impute' all missing data beforehand using external
% software to hard-call or define a distribution over the possible missing
% nucleotides and use that in the determination of path probabilities. Another,
% which we carry out within \texttt{sc2ts}, is to assume sites with missing
% nucleotides are uninformative to estimating the most likely path through the
% data which gives rise to the focal sequence. If we are able to estimate the
% most likely copying path among these paths, a reasonable imputation method is
% to fill in the missing data in the focal sequence by copying it from the
% relevant positions in the most likely path. To ensure that missing data is
% not considered in determining the most likely copying path, we set the
% emission probability from any state $(A,C,G,U,-)$ to `unknown' as 1 (though
% we could have chosen any constant). As a result, emissions of unknown
% nucleotides will not contribute to differences in path probabilities. Note
% that by iteratively imputing missing data in focal sequences in this way, the
% reference panel being used for any newly observed focal sequence is
% guaranteed to be devoid of missingness.

Genetic data for SARS-CoV-2 contains substantial amounts of missingness,
(nucleotides coded as `unknown' or masked out
as described in Section~\ref{sec:data_preprocessing})
and it is important to account for this missinging in a systematic way.
In \texttt{sc2ts} we automatically impute missing data as samples are added
into the ARG, using the LS model.
To do this, we assume that sites with missing nucleotides are uninformative
to the path probability, by setting the emission probability from any
state $(A,C,G,U,-)$ to `unknown' equal to 1 (though we could have chosen any constant).
As a result, emissions of unknown nucleotides will not contribute to
differences in path probabilities.
Once the most likely copying path is determined, we then attach the
sample to the ARG (see Figure~\ref{fig:overview_sc2ts} and following subsections).
For each newly attached sample we encode its nucleotide sequence by
recording mutations where it differs from the nucleotide sequence
of its parental node, but (importantly) ignoring sites with missing data
in the new sample.
Thus, once the newly added sequence has been attached to the ARG
any missing data is imputed from its parental node. There is therefore
no missing data in the ARG; all missing bases are ``hard called''
at attachment time.
Note that this approach is equivalent to
using state-of-the-art imputation
methods~\citep[e.g.][]{Browning2018-nk,Delaneau2019-wl} with a reference
panel consisting of all sequences in the ARG,
since these methods are also based on the LS model.
% Could say more here, but let's leave it.
Evaluating the accuracy of missing data imputation
using \texttt{sc2ts} is an important facet of future work.
